Spatial Reasoning and Data Displays

### 2015-06-26 Paper/InfoVis2015Revision/ResponsesToReviewers.Rmd

---
title: "Summary of Changes"
author: "Susan VanderPlas & Heike Hofmann"
output: word_document
---

We would like to thank the reviewers for the detailed comments. We hope to have addressed all of the questions.

We are detailing our response below:

# Major Concerns
1. **Are all the experimental design choices properly motivated?**    
**(refer to the first concern of R1).**

*We have extended the discussion of the choice of cognitive tests; the purpose of each of the cognitive tests is described in section 2.2. In summary, the paper-folding, card rotation, and figure classification tasks are part of the kit of factor-referenced cognitive tests (and are thus extremely well-studied); additionally, we included a visual searching task (VST) because of its similarity of form to statistical lineups and its' compatibility with the written form of the other cognitive tests and lineup tests.*    
*As mentioned in the paper, the paper folding and card rotation tests have been used in many previous studies of statistical graphics and data displays. The figure-classification task examines inductive reasoning, which is another component of the lineup task (determining the dimension on which the lineup components differ, then applying that to answer the question successfully).*
*XXX hypotheses about the relationship with lineups? XXX*

@@

2. **Are the experimental procedure sufficiently detailed?**    
**(refer to the second concern of R1 and )**    
*We have added additional detail about the experimental procedures, and have included the testing materials as supplementary documents (with the factor-referenced kit tests redacted due to copyright issues; in these cases only the instruction page is shown; however, there are various sources online, that -illegally- provide access to parts of the actual test sets, e.g. http://spatiallearning.org/resource-info/Spatial_Ability_Tests/Paper_Folding_Test.pdf ). The testing documents are comprehensive, and oral instructions mirror the instructions on the introduction page of each task (students are familiar with this format as it is commonly used on standardized tests, such as the SAT and ACT).*

### 2015-06-26 Paper/InfoVis2015Revision/TestingVisualAptitude.Rnw
couple of changes ...

%% Abstract section.
\abstract{
Graphics convey numerical information very efficiently, but rely on a different set of mental processes than tabular displays. This study examines the demographic characteristics and visual skills associated with perception of graphical lineups. We conclude that lineups are essentially a classification test in a visual domain, and that performance on the lineup protocol is associated with general aptitude, rather than specific tasks such as card rotation and spatial manipulation. We also examine the possibility that specific graphical tasks may be associated with certain visual skills and conclude that more research is necessary to understand which visual skills are required in order to understand certain plot types. 
} % end of abstract

@@

% \hhnote{XXX Still missing: we need to extend the discussion a little bit that we want to extract the common core from lineups rather than investigate the relationship between individual designs and the  tasks. I believe, that the lineup commonalities are stronger than the design differences - but we might need to get some statistical backup for that. ... one way to do that would be to do a PCA on the three lineups and investigate whether anything beyond the first dimension is significant. I don't think so.  }

% \hhnote{XXX we should add: the lineups come from three different lineup experiments that discuss and compare the design choices (cite papers). This is NOT what we are doing here. Regardless of design choices, we want to find out, what visual skills are necessary for dealing with lineups. }

@@

We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test.

The VST measures visual search speed~\cite{goldstein1973validity}, the paper folding and card rotation tests measures spatial manipulation ability, and the figure classification test measures inductive reasoning~\cite{ekstrom1976manual}; 
all of these skills are at least peripherally recruited during the lineup task, but some may dominate in predicting performance on the lineup task.

>>>>

%% Abstract section.
\abstract{
Graphics convey numerical information very efficiently, but rely on a different set of mental processes than tabular displays. here, we present a study relating demographic characteristics and visual skills to perception of graphical lineups. We conclude that lineups are essentially a classification test in a visual domain, and that performance on the lineup protocol is associated with general aptitude, rather than specific tasks such as card rotation and spatial manipulation. We also examine the possibility that specific graphical tasks may be associated with certain visual skills and conclude that more research is necessary to understand which visual skills are required in order to understand certain plot types. 
} % end of abstract

@@

@@

We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test. \newtext{The last three tests are part of the Kit of Factor-Referenced Cognitive Tests~}\cite{educational1963kit, ekstrom1976manual}. \newtext{These tests measure three separate dimensions of visuospatial ability and have been virtually unchanged in use  since the 1960s, and are thus extremely well-studied. }

The VST measures visual search speed~\cite{goldstein1973validity}; \newtext{we considered visual search as the basis of a lineup evaluation and included this test mainly to identify potential problems in the results, if participants were unable to perform this task.} 
Both  card rotation and paper folding tests measure spatial manipulation ability in \newtext{two- and three dimensions}, while the figure classification test measures inductive reasoning~\cite{ekstrom1976manual}; 
\newtext{we hypothesized that} all of these skills are %at least peripherally
recruited during the lineup task, but \newtext{are interested which of these dominate} in predicting performance on the lineup task.

### 2015-06-26 Paper/InfoVis2015Revision/TestingVisualAptitude.tex
couple of changes ...

% \hhnote{XXX Still missing: we need to extend the discussion a little bit that we want to extract the common core from lineups rather than investigate the relationship between individual designs and the  tasks. I believe, that the lineup commonalities are stronger than the design differences - but we might need to get some statistical backup for that. ... one way to do that would be to do a PCA on the three lineups and investigate whether anything beyond the first dimension is significant. I don't think so.  }

% \hhnote{XXX we should add: the lineups come from three different lineup experiments that discuss and compare the design choices (cite papers). This is NOT what we are doing here. Regardless of design choices, we want to find out, what visual skills are necessary for dealing with lineups. }

\label{sec:introduction}Data displays provide quick summaries of data, models, and results, but not all displays are equally good, nor is any data display equally useful to all viewers. 
Graphics utilize higher-bandwidth visual pathways to encode information~~\cite{baddeley1974working}, allowing viewers to quickly and intuitively relate multiple dimensions of numerical quantities.
Well-designed graphics emphasize and present important features of the data while minimizing  features of lesser importance, guiding the viewer towards conclusions that are meaningful in context and supported by the data while maximizing the information encoded in working memory. Under this framework, well-designed graphics reduce memory load and make more cognitive resources available for other tasks (such as drawing conclusions from the data), at the cost of depending on certain visuospatial reasoning abilities.

Many theories of graphical learning center around the difference between visual and verbal processing: the dual-coding theory emphasizes the utility of complementary information in both domains, while the visual argument hypothesis emphasizes that graphics are more efficient tools for providing data with spatial, temporal, or other implicit ordering, because the spatial dimension can be represented graphically in a more natural manner~~\cite{vekiri2002value}. Both of these theories suggest spatial ability impacts a viewer's use of graphics, because spatial ability either influences cognitive resource allocation or affects the processing of spatial relationships between graphical elements. In addition, previous investigations into graphical learning and spatial ability have found relationships between spatial ability and the ability to read information from graphs~~\cite{lowrie2007solving}. 
However,  mathematical ability, not spatial ability, was shown~~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph. 
Spatial ability becomes more important when more complicated graphical displays are used in comparison tasks: the lower performance of individuals with low spatial ability on tests utilizing diagrams and graphs is attributed ~~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. It is theorized that graphics are a form of ``external cognition"~~\cite{scaife1996external} that guide, constrain, and facilitate cognitive behavior~~\cite{zhang1997nature}.

Here, we want to investigate the extent to which observers make use of spatial ability in their evaluation of and reasoning about statistical graphics. As a quantitative measure of how successful an evaluation is, we are making use of the lineup protocol ~~\cite{buja2009statistical, wickham2010graphical, majumder2013validation}.

``Lineups" were introduced as a tool to evaluate the statistical significance of a graphical finding~\cite{wickham2010graphical}. Lineups are also useful in assessing the effectiveness of different graphical displays~\cite{hofmann2012graphical, loy:2015}. Like their police counterpart, lineups consist of several distractor plots (of randomly generated data) and one target (the data plot). Figure~\ref{fig:lineup}(a) shows a sample lineup of boxplots; participants are expected to identify the most different among the plots shown. In this example, sub-plot 4 is the target  because of the noticeably different locations of the two boxplot medians.

@@

We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test. 
The VST measures visual search speed~\cite{goldstein1973validity}, the paper folding and card rotation tests measures spatial manipulation ability, and the figure classification test measures inductive reasoning~\cite{ekstrom1976manual}; 
all of these skills are at least peripherally recruited during the lineup task, but some may dominate in predicting performance on the lineup task.

@@

% latex table generated in R 3.2.1 by xtable 1.7-1 package
% Thu Jun 25 09:34:25 2015
\begin{table}[htb]
\centering
\caption{Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\label{tab:PCAvariance4}} 
{\footnotesize
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Standard deviation & 1.61 & 0.81 & 0.73 & 0.49 \\ 
  Proportion of Variance & 0.64 & 0.16 & 0.13 & 0.06 \\ 
  Cumulative Proportion & 0.64 & 0.81 & 0.94 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

@@

% latex table generated in R 3.2.1 by xtable 1.7-1 package
% Thu Jun 25 09:34:25 2015
\begin{table}[htb]
\centering
\caption{Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\label{tab:PCArotation4}} 
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Card Rot & 0.55 & -0.19 & -0.38 & 0.72 \\ 
  Fig Class & 0.46 & 0.58 & 0.66 & 0.14 \\ 
  Folding & 0.52 & 0.33 & -0.53 & -0.59 \\ 
  VST & 0.46 & -0.72 & 0.38 & -0.34 \\ 
   \hline
\end{tabular}
\end{table}

@@

Next, we add the lineup results to our principal components analysis. The results of the principal component analysis of the four cognitive tests and lineup results are similar to the results of the cognitive tests alone; this suggests that lineups require at least some of the skills measured by the other cognitive tests. Table~\ref{tab:PCAvariance5} shows the importance of each principal component; from the distribution of the variance components, we see that the lineup test spans an additional dimension within the space of the four established tests. 
% latex table generated in R 3.2.1 by xtable 1.7-1 package
% Thu Jun 25 09:34:26 2015
\begin{table}[ht]
\centering
\caption{Importance of principal components, analyzing all five tests.\label{tab:PCAvariance5}} 
{\footnotesize
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
Standard deviation & 1.73 & 0.84 & 0.75 & 0.70 & 0.48 \\ 
  Proportion of Variance & 0.60 & 0.14 & 0.11 & 0.10 & 0.05 \\ 
  Cumulative Proportion & 0.60 & 0.74 & 0.85 & 0.95 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

% latex table generated in R 3.2.1 by xtable 1.7-1 package
% Thu Jun 25 09:34:26 2015
\begin{table}[ht]
\centering
\caption{PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \label{tab:PCArotation5}} 
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
lineup & 0.42 & 0.49 & -0.46 & 0.60 & -0.10 \\ 
  card.rot & 0.50 & -0.30 & 0.28 & 0.23 & 0.73 \\ 
  fig.class & 0.43 & 0.45 & -0.15 & -0.75 & 0.18 \\ 
  folding & 0.47 & 0.07 & 0.68 & 0.04 & -0.56 \\ 
  vis.search & 0.41 & -0.69 & -0.48 & -0.15 & -0.33 \\ 
   \hline
\end{tabular}
\end{table}

@@

% latex table generated in R 3.2.1 by xtable 1.7-1 package
% Thu Jun 25 09:34:26 2015
\begin{table}[ht]
\centering
\caption{Estimates and significances of a linear model of lineup scores.} 
\label{tab:m1}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 14.1192 & 1.9149 & 7.37 & 0.0000 \\ 
  PC1 & 1.7672 & 0.5230 & 3.38 & 0.0018 \\ 
  MATH & 2.1246 & 0.8732 & 2.43 & 0.0202 \\ 
   \hline
\end{tabular}
\end{table}

%\afterpage{\clearpage}
%
\subsection{Visual Aptitude for Specific Lineup Tasks}
Figure~\ref{fig:lineuptaskmatrix} shows the correlations between the three lineup tasks and the figure classification, card rotation, and paper folding tasks. 
%The visual search task is only slightly correlated with the three lineup tasks and is therefore omitted from this figure.
Performance on lineup tasks 1 and 2, which dealt with the distributuion of two groups of numerical data, is most strongly correlated with the performance on the figure classification task, which measures general reasoning ability.
Performance on lineup task 3, which investigated the potential to visually identify nonnormality in residual QQ-plots, is more associated with the card rotation and paper folding tests, which measure visuospatial ability. This suggests that certain lineup tasks may require more visual ability than others; in the case of the QQ-lineups a successful evaluation  needed  participants to mentally rotate plots to compare vertical distances, requiring more mental manipulation than the first two lineup tasks, which required participants to examine categorical data.

>>>>

\label{sec:introduction}Data displays provide quick summaries of data, models, and results, but not all displays are equally good, nor is any data display equally useful to all viewers. 
Graphics utilize higher-bandwidth visual pathways to encode information~\cite{baddeley1974working}, allowing viewers to quickly and intuitively relate multiple dimensions of numerical quantities.
Well-designed graphics emphasize and present important features of the data while minimizing  features of lesser importance, guiding the viewer towards conclusions that are meaningful in context and supported by the data while maximizing the information encoded in working memory. Under this framework, well-designed graphics reduce memory load and make more cognitive resources available for other tasks (such as drawing conclusions from the data), at the cost of depending on certain visuospatial reasoning abilities.

Many theories of graphical learning center around the difference between visual and verbal processing: the dual-coding theory emphasizes the utility of complementary information in both domains, while the visual argument hypothesis emphasizes that graphics are more efficient tools for providing data with spatial, temporal, or other implicit ordering, because the spatial dimension can be represented graphically in a more natural manner~\cite{vekiri2002value}. Both of these theories suggest spatial ability impacts a viewer's use of graphics, because spatial ability either influences cognitive resource allocation or affects the processing of spatial relationships between graphical elements. In addition, previous investigations into graphical learning and spatial ability have found relationships between spatial ability and the ability to read information from graphs~\cite{lowrie2007solving}. 
However,  mathematical ability, not spatial ability, was shown~\cite{shah1995conceptual} to be associated with accuracy on a simple two-dimensional line graph. 
Spatial ability becomes more important when more complicated graphical displays are used in comparison tasks: the lower performance of individuals with low spatial ability on tests utilizing diagrams and graphs is attributed~\cite{mayer1994whom} to the fact that more cognitive resources are required to process the visual stimuli, which leaves fewer resources to make connections and draw conclusions from those stimuli. It is theorized that graphics are a form of ``external cognition"~\cite{scaife1996external} that guide, constrain, and facilitate cognitive behavior~\cite{zhang1997nature}.

Here, we want to investigate the extent to which observers make use of spatial ability in their evaluation of and reasoning about statistical graphics. As a quantitative measure of how successful an evaluation is, we are making use of the lineup protocol~\cite{buja2009statistical, wickham2010graphical, majumder2013validation}.

``Lineups" were introduced as a tool to evaluate the statistical significance of a graphical finding~\cite{wickham2010graphical}. Lineups are also useful in assessing the effectiveness of alternative graphical displays~\cite{hofmann2012graphical, loy:2015}. Like their police counterpart, graphical lineups consist of several distractor plots (of randomly generated data) and one target (the data plot).  Figure~\ref{fig:lineup}(a) shows a sample lineup of boxplots; in this example, sub-plot 4 shows the target data (recognizable by the shift between the two boxplot medians).

@@

We compare lineup performace to the visual search task (VST), paper folding test, card rotation test, and figure classification test. \newtext{The last three tests are part of the Kit of Factor-Referenced Cognitive Tests~}\cite{educational1963kit, ekstrom1976manual}. \newtext{These tests measure three separate dimensions of visuospatial ability and have been virtually unchanged in use  since the 1960s, and are thus extremely well-studied. }

The VST measures visual search speed~\cite{goldstein1973validity}; \newtext{we considered visual search as the basis of a lineup evaluation and included this test mainly to identify potential problems in the results, if participants were unable to perform this task.} 
Both  card rotation and paper folding tests measure spatial manipulation ability in \newtext{two- and three dimensions}, while the figure classification test measures inductive reasoning~\cite{ekstrom1976manual}; 
\newtext{we hypothesized that} all of these skills are %at least peripherally
recruited during the lineup task, but \newtext{are interested which of these dominate} in predicting performance on the lineup task.

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Fri Jun 26 14:46:58 2015
\begin{table}[htb]
\centering
\caption{Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\label{tab:PCAvariance4}} 
{\footnotesize
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Standard deviation & 1.61 & 0.81 & 0.73 & 0.49 \\ 
  Proportion of Variance & 0.64 & 0.16 & 0.13 & 0.06 \\ 
  Cumulative Proportion & 0.64 & 0.81 & 0.94 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Fri Jun 26 14:46:59 2015
\begin{table}[htb]
\centering
\caption{Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\label{tab:PCArotation4}} 
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Card Rot & 0.55 & -0.19 & -0.38 & 0.72 \\ 
  Fig Class & 0.46 & 0.58 & 0.66 & 0.14 \\ 
  Folding & 0.52 & 0.33 & -0.53 & -0.59 \\ 
  VST & 0.46 & -0.72 & 0.38 & -0.34 \\ 
   \hline
\end{tabular}
\end{table}

@@

Next, we add the lineup results to our principal components analysis. The results of the principal component analysis of the four cognitive tests and lineup results are similar to the results of the cognitive tests alone; this suggests that lineups require at least some of the skills measured by the other cognitive tests. Table~\ref{tab:PCAvariance5} shows the importance of each principal component; from the distribution of the variance components, we see that the lineup test spans an additional dimension within the space of the four established tests. 
% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Fri Jun 26 14:46:59 2015
\begin{table}[ht]
\centering
\caption{Importance of principal components, analyzing all five tests.\label{tab:PCAvariance5}} 
{\footnotesize
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
Standard deviation & 1.73 & 0.84 & 0.75 & 0.70 & 0.48 \\ 
  Proportion of Variance & 0.60 & 0.14 & 0.11 & 0.10 & 0.05 \\ 
  Cumulative Proportion & 0.60 & 0.74 & 0.85 & 0.95 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Fri Jun 26 14:46:59 2015
\begin{table}[ht]
\centering
\caption{PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \label{tab:PCArotation5}} 
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
lineup & 0.42 & 0.49 & -0.46 & 0.60 & -0.10 \\ 
  card.rot & 0.50 & -0.30 & 0.28 & 0.23 & 0.73 \\ 
  fig.class & 0.43 & 0.45 & -0.15 & -0.75 & 0.18 \\ 
  folding & 0.47 & 0.07 & 0.68 & 0.04 & -0.56 \\ 
  vis.search & 0.41 & -0.69 & -0.48 & -0.15 & -0.33 \\ 
   \hline
\end{tabular}
\end{table}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Fri Jun 26 14:46:59 2015
\begin{table}[ht]
\centering
\caption{Estimates and significances of a linear model of lineup scores.} 
\label{tab:m1}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 14.1192 & 1.9149 & 7.37 & 0.0000 \\ 
  PC1 & 1.7672 & 0.5230 & 3.38 & 0.0018 \\ 
  MATH & 2.1246 & 0.8732 & 2.43 & 0.0202 \\ 
   \hline
\end{tabular}
\end{table}

%\afterpage{\clearpage}
%
\subsection{Visual Aptitude for Specific Lineup Tasks}
Figure~\ref{fig:lineuptaskmatrix} shows the correlations between the three lineup tasks and the figure classification, card rotation, and paper folding tasks. 
%The visual search task is only slightly correlated with the three lineup tasks and is therefore omitted from this figure.

Performance on lineup tasks 1 and 2, which dealt with the distributuion of two groups of numerical data, is most strongly correlated with the performance on the figure classification task, which measures general reasoning ability.
Performance on lineup task 3, which investigated the potential to visually identify nonnormality in residual QQ-plots, is more associated with the card rotation and paper folding tests, which measure visuospatial ability. This suggests that certain lineup tasks may require more visual ability than others; in the case of the QQ-lineups a successful evaluation  needed  participants to mentally rotate plots to compare vertical distances, requiring more mental manipulation than the first two lineup tasks, which asked from participants to examine categorical data.
